---
title: "A latent trajectory analysis in glycemic control and diabetic retinopathy screening coverage 2011-2023: A municipality-level study"
author:
  - name: Rolando Silva-Jorquera
    orcid: 0000-0002-1244-8350
    email: rolando.silva@pucv.cl
    affiliations:
      - name: University College London, Institute for Global Health, London, United Kingdom
  - name: Kasim Allel
    orcid: 0000-0002-2144-7181
    email: kasim.allel1@lshtm.ac.uk
    affiliations:
      - name: University of Oxford, Nuffield Department of Population Health, Oxford, United Kingdom
  - name: Paul Nderitu
    orcid: 0000-0002-8571-0751
    email: paul.nderitu@kcl.ac.uk
    affiliations:
      - name: King’s College London, Section of Ophthalmology, Faculty of Life Sciences and Medicine, London, United Kingdom
      - name: King’s College London, Department of Ophthalmology, King’s Ophthalmology Research Unit (KORU), London, United Kingdom
  - name: Hassan Haghparast Bidgoli
    orcid: 0000-0001-6365-2944
    email: h.haghparast-bidgoli@ucl.ac.uk
    affiliations:
      - name: University College London, Institute for Global Health, London, United Kingdom
  - name: Claudio Zett
    orcid: 0000-0003-1839-0927
    email: claudio.zett@pucv.cl
    affiliations:
      - name: Pontificia Universidad Católica de Valparaíso, Valparaíso, Chile
format:
  html:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    theme: readable
  pdf:
    number-sections: true
    colorlinks: true
    keeptex: true
    include-in-header: 
      text: |
        \usepackage{booktabs}
        \usepackage{siunitx}
        \newcolumntype{d}{S[
            input-open-uncertainty=,
            input-close-uncertainty=,
            parse-numbers = false,
            table-align-text-pre=false,
            table-align-text-post=false
         ]}
  docx: default
prefer-html: true
date: 'last-modified'
#date-format: '[This version:] MMMM D, YYYY [<br>(First version: February  16, 2023)]'
#abstract: "**Background:** Diabetes management in Chile varies from that in other countries. The Cardiovascular Program (PSCV) aims to control diabetes and retinopathy screening; however, this indicator has not been consistently monitored. Latent class mixture modeling was used to investigate the factors influencing membership in different subpopulations. **Aim**: This study analyzed PSCV data from 2011 to 2023 to identify unique trajectories in diabetes and diabetic retinopathy screening coverage (DMC and DRSC) among municipalities. The characteristics of deprivation were also examined. The analyses were conducted at the ecological level to identify municipalities with coverage differing from the national average or target. **Methods:** This study obtained data from the Statistical and Health Information Department on individuals with T2DM covered by the public health scheme, aged 15 years or above, and eligible for an annual eye examination in all 346 municipalities in Chile between 2011 and 2023. To have stronger analysis we remove the clasified municiplities by size of DM population, and the the lower quintil was removed.A linear Latent Class Mixed Model was used to derive trajectory classes for DRSC and DGCC, with six clusters identified based on clinical plausibility, model fit, and entropy criteria. Logistic regression was used to determine the impact of municipality characteristics on trajectory-class membership. This study focuses on interpreting the shape of clusters from a theoretical perspective. **Results:**  For each of the six models, the fit criteria were evaluated and the proportions in each trajectory class were calculated, along with the estimated intercepts and slopes for each class, with random effects across municipalities. The one-class model tested was a simple linear model with years predicting coverage and cluster membership, with intercepts representing baseline coverage and slopes indicating average annual change. The one-class model for DGC showed an average coverage of 45.10% at baseline with a small but significant negative slope, whereas the one-class model for DRS showed an average coverage of 28.45% with an annual increase of 1.59. Further analysis will focus on the DRSC's class membership, and the two-class model was preferred for DRSC over the one- and three-class models. In 2011, Class 1 (74.32% of municipalities) had an average coverage of 22.88% and experienced a gradual annual increase of 1.41% until 2023, whereas Class 2 (25.68%) had an average of 44.03% in 2011 and experienced a rapid annual increase of 2.07% until 2023. The unadjusted associations of socioeconomic index predictor factors with municipality DRSC trajectory provided an OR estimate of 0.531 (95% CI, 0.387–0.716), suggesting that as deprivation levels rise, the probability of achieving higher coverage (Class 2) increases. **Conclusion:** The DGCC in LAs in Chile appears to follow only a single trajectory, whereas the DRSC follows two slowly growing differing trajectories. Counterintuitively, the trajectories that followed did not appear to be influenced by deprivation. The group of municipalities following a higher DRSC trajectory is potentially worthy of further study to identify whether there are common local policies or practices that may have contributed to their positive deviance from national trends towards an increase in DRSC."
bibliography: references.bib
csl: vancouver-superscript.csl
tbl-cap-location: top
number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
editor:
  mode: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
# Use cache = TRUE if you want to speed up compilation

knitr::opts_knit$set(output.format = "html")  # Set to "html" for HTML output


# A function to allow for showing some of the inline code
rinline <- function(code){
  html <- '<code  class="r">``` `r CODE` ```</code>'
  sub("CODE", code, html)
}

```

```{r}
#| eval: true
#| echo: false
#| message: false

library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggtext)
library(stringr)
library(forcats)
library(Cairo)
library(extrafont)
library(hrbrthemes)
library(directlabels)
library(ggrepel)
library(readxl)
library(extrafont)
library(scales)
library(ggrepel)
library(ggsci)
library(lcmm)
library(MixAll)
require(kml)
require(traj)
require(lcmm)
require(lmerTest)
require(plyr)
require(tidyr)
require(psych)
require(fpc)
require(mclust)
require(rcompanion)
library(gridExtra)
library(tidyLPA)
library(MASS)
library(broom)
library(skimr)
library(gtExtras)
library(psych)
library(pander)
library(tibble)
library(BayesFactor)
library(modelsummary)
library(gt)
library(gtsummary)
library(survival)
library(xtable)
library(skimr)

options(scipen=10000000)


```

**ABSTRACT**

**Purpose:** Diabetic retinopathy screening (DRS) and glycemic control are crucial for managing type 2 diabetes, yet access and outcomes vary regionally. This study uses latent class mixture modeling (LCMM) on longitudinal data from Chilean municipalities to identify trajectory patterns in type 2 diabetes management and DRS coverage and to explore their associations with municipal demographics and deprivation, highlighting subpopulations with unique healthcare needs.

**Methods:** We analyzed data from 2011 to 2023 for adults aged ≥15 years with type 2 diabetes from Chile’s public health system. This study focused on Diabetic Glycemic Control Coverage (DGCC) and DRS Coverage (DRSC). Data were obtained from the Chilean Public Health Platform, excluding smaller municipalities. Associations between coverage trajectories and municipal factors, including deprivation, urbanicity (rural, mixed, and urban), and geographical location (northern, central, and southern zones), were explored.

**Results:** The analysis included 300 municipalities (2011-2023) and 292 municipalities (2011-2019), examining DGCC and DRSC. Baseline DGCC was 46.7% (2011-2023) and 45.1% (2011-2019), with annual declines of -1.03% (p \< 0.001) and -0.42% (p \< 0.001), respectively. DRSC increased from 28.5% (2011-2019) to 31.3% (2011-2023), with annual increases of +1.59% (p \< 0.001) and +0.5% (p \< 0.001). DGCC followed a one-class trajectory, whereas DRSC followed two classes. Class 1 (lower coverage) had baselines of 24.8% (2011-2023) and 23.05% (2011-2019), with annual changes of +0.4% (p = 0.068) and +1.38% (p \< 0.001), respectively. Class 2 (higher coverage) had baselines of 43.7% (2011-2023) and 43.58% (2011-2019), with annual changes of +0.69% (p = 0.073) and +2.16% (p = 0.002). Deprivation (lower ISED scores) was linked to lower DRSC coverage: OR 0.66 (2011-2023, 95% CI: 0.50, 0.85) and 0.53 (2011-2019, 95% CI: 0.39, 0.73). Rural areas had higher odds for higher coverage: OR 1.79 (2011-2023, 95% CI: 0.99, 3.21) and OR 2.73 (2011-2019, 95% CI: 1.41, 5.30). Urban areas had lower odds: OR 0.39 (2011-2023, 95% CI: 0.17, 0.84) and OR 0.47 (2011-2019, 95% CI: 0.19, 1.19). The southern region showed higher odds for higher coverage: OR 1.02 (2011-2023, 95% CI: 0.42, 2.51) and OR 1.84 (2011-2019, 95% CI: 1.11, 3.08).

**Conclusions:** DGCC in Chilean municipalities followed a single trajectory, while DRSC exhibited two distinct trajectories, one with lower coverage and one with higher coverage. Lower deprivation levels were associated with a higher likelihood of having a higher DRSC trajectory. Further exploration of municipalities with higher DRSC coverage may provide insights into the factors contributing to these patterns, potentially highlighting local policies or practices that could be effective. Future research should investigate the factors influencing DRSC coverage trajectories over time and evaluate strategies to improve DGCC and DRSC coverage.

**Keywords**: Latent class trajectories analysis, diabetic retinopathy screening trajectories, diabetes mellitus

# Introduction

<!--

Previous intro 1 - 

The rise in the prevalence of diabetes mellitus (DM), particularly type 2 DM (T2DM),[@idf_2021] has placed strain on health systems, particularly in the management of vision-threatening issues resulting from poor metabolic control. Diabetic retinopathy (DR) is the leading cause of avoidable blindness and vision impairment in working-age adults.[@burton_lancet_2021] The consequences of inadequate diabetes management, coupled with the lack of early DR detection and treatment, not only cause social and emotional distress,[@fenwick_social_2012] but also limit participation in daily activities and job opportunities.[@cooper_psychological_2020] This, in turn, leads to significant economic repercussions[@pdb52_2019] and has a profound impact on the quality of life,[@trento_quality_2013] presenting substantial public health concerns. Therefore, a robust diabetes management system, including regular DR screening (DRS), is crucial to address this pressing issue. However, the effectiveness of diabetes management and efforts to prevent sight loss vary significantly at the local level in Chile[@SILVA-JORQUERA2021; @guerrero-núñez2017] and other countries[@elek2021] and are influenced by factors such as deprivation, geographical location and level of urbanicity, and may also reflect differences in healthcare service policy and provision related to DM.[@guerrero-núñez2017; @elek2021]

In line with the targets and commitments of the 2030 Agenda for Sustainable Development, the World Health Organization has called on health systems to accelerate actions to improve DM detection and quality of care, including achieving good glycemic control (HbA1c \< 7%) in 80% of the diagnosed population.[@who] On the other hand, DR no longer being the leading cause of blindness among working-age adult in England, after a implementation of a systematic organised DRS nationwide screening program[@liew_comparison_2014] This program, has achieved 80% of the people diagnosed with DM annually.[@Scanlon_2017] Overall, in line with the WHO call, health systems should aims to improve the coverage of people achieving a HbA1C \<7% as a 80%, and improve early detection and time treatment of people sight loss threatened by DM.

The Cardiovascular Health Program (CHP) in Chile aims to monitor trends in individuals with type 2 diabetes mellitus (T2DM) and has established guidelines for monitoring glycated hemoglobin (HbA1c) levels and annual retinopathy (DR) screening for individuals aged 15 years or older.[@minsal2017] In 2021, an analysis was conducted to investigate long-term trends in DRS coverage (DRSC).[@SILVA-JORQUERA2021] The analysis revealed significant heterogeneity and considerable variation in longitudinal trends. Such findings raise the question whether there are distinct trajectories within sub-populations of municipalities about the proportion of individuals achieving HbA1c levels of \< 7% and undergoing annual screening for DR among those with T2DM over time. By identifying potentially different pattern trajectories in these indicators, decision-making and resource allocation can be informed to address the burden of T2DM and vision loss within the health system, in line with the World Health Organization's (WHO) call[@who] and existing evidence for improving this coverage.[@Scanlon_2017]

Latent class mixture modeling (LCMM) is an unsupervised machine learning and certainly a data reduction technique used to identify or uncover sub-populations within a larger population based on observed data.[@berlin_introduction_2014] LCMM has been particularly useful in psychiatric and psychological research, where it can be employed to study the heterogeneity of complex phenomena and address research questions related to treatment effects.[@miettunen2016] Furthermore, its applicability extends to a broader range of topics, including epidemiology.[@viner2018] Therefore, such techniques might assist us in addressing this practical problem, which, to the best of our knowledge, has not yet been studied.

Consequently, this study aimed to fill this research gap by determining whether there are specific groups or sub-populations of municipalities with unique trajectory patterns in achieving HbA1c targets of \<7% and DRSC by analysing data at the population level. Additionally, this study sought to investigate the relationship between area-level demographic and deprivation characteristics with these identified trajectories. Although these analyses were conducted at the ecological level, the results may be useful for identifying groups of municipalities that perform better or worse than the national average, providing valuable information for further research.

Previous intro 2 - 

The global rise in the prevalence of diabetes mellitus (DM), particularly type 2 DM (T2DM), has placed a significant strain on health systems worldwide [@idf_2021]. This burden is particularly acute in managing vision-threatening complications arising from poor glycemic control. Diabetic retinopathy (DR), a leading cause of avoidable blindness and vision impairment in working-age adults [@burton_lancet_2021], poses a major public health concern. Inadequate diabetes management, coupled with a lack of early DR detection and treatment, can have devastating consequences beyond vision loss. Studies suggest it can lead to social and emotional distress [@fenwick_social_2012], limit participation in daily activities and job opportunities [@cooper_psychological_2020], and have significant economic repercussions [@pdb52_2019]. Ultimately, these factors can profoundly impact the quality of life for individuals with T2DM [@trento_quality_2013]. Therefore, a robust diabetes management system, including regular DR screening (DRS), is crucial to address this pressing issue.

However, the effectiveness of these strategies varies significantly at the local level in Chile, as evidenced by studies highlighting inequalities in DRS coverage across municipalities [@SILVA-JORQUERA2021; @guerrero-núñez2017]. Similar findings have been reported in other countries [@elek2021]. These inequalities are likely influenced by factors such as deprivation, geographical location and level of urbanicity, and may also reflect differences in healthcare service policy and provision related to DM management [@guerrero-núñez2017; @elek2021].

Given the substantial public health burden of T2DM and its vision-threatening complications, particularly DR, effective management strategies are crucial. While national guidelines in Chile establish protocols for HbA1c monitoring and annual DRS [@minsal2017], prior studies highlight significant inequalities in achieving these targets across municipalities [@SILVA-JORQUERA2021]. These findings suggest potential heterogeneity within the population regarding glycemic control and DRS coverage. Understanding this heterogeneity is vital for developing targeted interventions to address these inequalities and improve overall diabetes care at the local level.

Traditional statistical approaches often assume homogeneity within the population under study. However, in the context of T2DM management, there might be distinct sub-populations with unique trajectories in achieving glycemic control and undergoing DRS. Latent class mixture modeling (LCMM) is an unsupervised machine learning technique that can identify these sub-populations based on observed characteristics [@berlin_introduction_2014; @miettunen2016]. By applying LCMM to longitudinal data on HbA1c levels and DRS coverage across Chilean municipalities, we aim to reveal unique trajectory patterns in achieving glycemic control and undergoing DRS.

Previous intro 3 -

Diabetes Mellitus (DM), a chronic metabolic disorder characterized by elevated blood glucose levels, affected 537 million people globally in 2021, with projections indicating a further increase to 783 million by 2045.[@idf_2021] Among these cases, 96% (95% CI 95.1--96.8) were attributed to type 2 DM (T2DM), making it the primary contributor to the overall DM burden at 95.4% (95% CI 94.9--95.9)[@ong_global_2023] One-third of the individuals living with DM are affected by diabetic retinopathy (DR),[@burton_lancet_2021] a prevalent complication of T2DM. DR is associated with longer DM duration, poorer glycemic control, and high blood pressure.[@yau_global_2012] Furthermore, DR is the leading cause of moderate-to-severe visual impairment, affecting 1.25% of cases, and accounts for 1.07% of blindness cases.[@burton_lancet_2021] DR it is the leading cause of sight loss worldwide, particularly among working-age adults.[@burton_lancet_2021]

The consequences of sight loss resulting from DM extend beyond individual well-being, encompassing social and emotional stress[@fenwick_social_2012], reduced participation in daily activities, and limited employment opportunities[@cooper_psychological_2020]. This significant public health issue not only impacts the quality of life[@trento_quality_2013] but also incurs substantial economic consequences[@pdb52_2019]. Therefore, an efficient DM management system, including regular DR screening (DRS), is essential for addressing this pressing problem.

For optimal DRS, transitioning from opportunistic to systematically organized DRS is crucial. Opportunistic screening relies on individuals seeking routine health care visits based on their personal circumstances. In contrast, systematically organized programs proactively identify at-risk populations for regular DRS, ensuring broad and consistent coverage. Achieving approximately 80% annual coverage, systematic screening, as exemplified by successful implementation in countries such as England and Wales, significantly reduces sight loss.[@Scanlon_2017] In England, diabetic retinopathy has no longer been the leading cause of blindness among working-age individuals since the implementation of a nationwide DRS program[@liew_comparison_2014], marking a strategic shift in the screening approach.

On the other hand, the extent of diabeteic glycemic control coverage (DGCG) and DRS coverage(DRSC) varies significantly at the global district level[@SILVA-JORQUERA2021; @guerrero-núñez2017, @elek2021]. Socioeconomic factors, coupled with local government policies, play a role in these variations[@guerrero-núñez2017; @elek2021], with pronounced disparities in nations marked by significant income inequality, such as Chile[@lynch2004]. Interestingly, Chile, which lacks a formal DRS programme, has an opportunistic rolling-out DRS. Its highest point in DRSC achieved at 36% in the last decade; less than a half from the 80% target.[@SILVA-JORQUERA2021] However, its specialized eye workforce and a growing eye health infrastructure, positioned to this country in an stage with a significant potential to move towards a systematic DRS approach, leveraging its existing resources and infrastructure to enhance DM management and improve outcomes.

In 2021, an analysis using health datasets at the ecological level from 2011 to 2019 highlighted substantial variations in DRSC when municipalities were grouped by 29 administrative healthcare services (e.g., the lowest group at 15% and the highest averaging 50%, a threefold difference)[@SILVA-JORQUERA2021].Unfortunately, Chile experienced a massive disruption due to COVID-19, mirroring global challenges in managing DM and DRS.[@ahmed2021] Cut DRSC by half and left over 500 thousand individuals with inadequate glycemic control with no annual DRS in 2020.[@silva-jorquera2023] However, there are a still insights to explore DGCC and DRSC to inform policy and practice for better desicion making and resource allocation, specially for planning a systematic DRS. The study prompted inquiries into the existence of different local practices and socioeconomic gradients at the municipality level, including geographical territories, potentially driving DRSC and DGCC.

This study examined the trajectory patterns in the DGCC and DRSC across geographic regions and municipalities in Chile between 2011-2023. We explored the potential impact of COVID-19 on trajectory patterns and the relationship between deprivation characteristics.

-->

<!-- The Rising Burden of Diabetes and Diabetic Retinopathy-->

The global increase in diabetes mellitus (DM), particularly type 2 DM (T2DM), has placed an immense burden on healthcare systems worldwide.[@idf_2021] This strain is particularly pronounced in the management of vision-threatening complications arising from inadequate glycemic control. Diabetic retinopathy (DR) remains a leading cause of avoidable blindness and vision impairment among working-age adults and poses a critical public health challenge.[@burton_lancet_2021] In response, the World Health Organization (WHO) has emphasized the importance of robust diabetes management and care quality, including maintaining glycated hemoglobin (HbA1c) levels below 7%.[@who] This target aligns with the success observed in England, where a national diabetic retinopathy screening (DRS) program has significantly reduced DR as a cause of blindness in adults.[@liew_comparison_2014]

<!--Inequalities in Diabetes Management and Screening Coverage-->

Despite these efforts, many individuals with T2DM still face social and emotional distress,[@fenwick_social_2012] limited daily functioning and job opportunities,[@cooper_psychological_2020] and significant economic consequences of unmanaged DR and diabetes.[@pdb52_2019] In turn, these impacts profoundly affect quality of life.[@trento_quality_2013] However, access to effective diabetes management and DRS varies significantly across local regions, as seen in Chile, where inequalities in DRS coverage (DRSC) exist among municipalities.[@SILVA-JORQUERA2021; @guerrero-núñez2017] Similar disparities have been documented in other countries, often influenced by socioeconomic factors, urbanicity, and geographic location,[@patel2022; @elek2021; @ravindranath2023; @davis2023; @rawlingschidi2024] which affect access to healthcare and policy implementation for diabetes management.[@guerrero-núñez2017; @elek2021]

<!--Understanding Sub-population Heterogeneity in the Context of Inequalities-->

Addressing these inequalities in Chile requires a clear understanding of the variations in DRS coverage across municipalities. Chile’s national guidelines established by the Cardiovascular Health Program call for regular HbA1c monitoring and annual DRS to prevent DR.[@minsal2017] However, there are considerable disparities in meeting these targets across different areas.[@SILVA-JORQUERA2021; @guerrero-núñez2017] Identifying subpopulations with unique healthcare needs and understanding how factors such as demographics and socioeconomic status affect their healthcare trajectories could inform targeted interventions. Such efforts may ultimately improve local-level diabetes management and access to DRS.

<!--Latent Class Mixture Modeling: A Novel Approach-->

Traditional statistical approaches may overlook these population-level variations, often assuming a uniform response to diabetes management across the study population. Distinct subpopulations with unique trajectories for achieving glycemic control and accessing DRS services are likely to exist. Latent class mixture modelling (LCMM) is an unsupervised machine learning technique that can identify these subpopulations based on observed characteristics.[@berlin_introduction_2014; @miettunen2016] However, to our knowledge, no prior studies have applied LCMM to explore subpopulation heterogeneity within the context of T2DM management and DRSC in Chile.

In this study, we applied LCMM to longitudinal data on T2DM management and DRS coverage in Chilean municipalities. Our aim was to reveal unique trajectory patterns in achieving glycemic control and access to the DRS. Additionally, we investigated how area-level demographics and deprivation characteristics correlate with these identified trajectories. Although conducted at an ecological level, our findings offer a novel approach to identifying municipalities that perform above or below the national average. These insights have the potential to guide future research and inform policy interventions to reduce DR and improve equity in diabetes management.

# Methods

<!-- We analyzed municipality-level data from 2011 to 2023, focusing on adults aged ≥ 15 years diagnosed with type 2 diabetes mellitus (T2DM) and covered by the public health scheme. Data were sourced from the Statistical and Health Information Department, which provides T2DM management metrics through the CHP.[@deis] We used two outcome indicators: the proportion of T2DM patients achieving target glycemic control (HbA1c \< 7%), referred to as diabetic glycemic control coverage (DGCC), and those undergoing annual diabetic retinopathy screening (DRS). DGCC represents the percentage of the T2DM population maintaining an HbA1c level below 7%, while DRS coverage (DRSC) indicates the percentage undergoing annual DRS. Numerators for both metrics were based on the population with glycemic control \< 7% for DGCC and those undergoing annual DRS for DRSC, with the denominator equal to the total T2DM population. Municipalities in the lowest population size quintile were excluded due to their limited T2DM population representation. Due to the incomplete availability of sex-specific data, integrated data were used. To test the associations between various socio-demiographics factors and class membership coverage, we considered deprivation, urbanicity, and geographical location.

For deprivation, we utilized the index of socioeconomic development (ISDE) score, developed in 2013 by the Chilean Observatory of Public Health (OCHISAP). This index evaluates local deprivation levels (ranging from 0 to 1, with lower scores indicating greater deprivation) and comprises three components: income, education, and housing and sanitation.[@gattini_comunas_2013]

Urbanicity was determined by the National Statistics Institute, following the methodological framework outlined in the National Policy of Rural Development.[@bergamini2020] Municipalities were categorized based on population density as predominantly rural (more than 50% of the population residing in low-density areas), mixed (25% to 49% in low-density areas), and predominantly urban (less than 25% in low-density areas).

For geographical location, we used the classification by the Environmental Tribunals of Chile, which subdivides the country into three zones: North, Centre, and South. This classification simplifies the geographical location of municipalities into three levels.

Thus, we analyzed the unadjusted and adjusted associations with deprivation (ISDE score), zone or macro-region (North, Centre, South), and urbanicity (classification as Rural, Mixed, or Urban).

## Statistical analyses

We employed a two-stage analysis. First, we used latent class mixture modeling (LCMM) to analyse DRSC and DRSC longitudinal data. LCMM uses probabilistic assignment to latent classes, which represent subgroups or trajectories and allows the identification of unique patterns within a given population[@berlin_introduction_2014]. We delineated trajectory classes for DRSC and DGCC from 2011 to 2023. We gradually expanded from one class cluster into six classes clusters. Models were evaluated according to their goodness-of-fit, including the Akaike Information Criteria (AIC), Bayesian Information Criteria (BIC), sample-adjusted Bayesian Information Criteria (SABIC), Consistent Akaike Information Criterion (CAIC), Approximate Weight of Evidence Criterion (AWE), Bayes Factor (BF), and correct model probability (cmP). We employed diagnostic tests, including entropy, smallest class count and size, minimun Average Latent Class Posterior Probability (ALCPP). The Vuong--Lo--Mendell--Rubin test assessed overall model fit, comparing a model with k-classes to a model with k-1 classes. The rejection of the k-1 class model in favor of the k-class model was based on a p-value threshold of p\<0.05. When evaluating models, we considered clinical plausibility, model fit, and diagnostic criteria.[@wickrama2021] We hypothesized that the intercept and slope represented random effects across the municipalities. We conducted a sensitivity analysis spanning the time between 2011 and 2019, mirroring our initial analysis to evaluate whether excluding the pandemic years (2019-2023) impacted DRSC and DRSC trajectory results. As LCMM deal with missing data no neccesary to have a complete data for each year in each municipality. However we have checked the data completeness over the time by comparing means using Welch's two sample t -test and verify whether data were balanced between selected trajectories.

Second, we examined the factors influencing membership across different trajectories. We performed logistic regression analysis to determine whether municipal-level socioeconomic development (measured by the ISDE), urbanicity, and macro-region were associated with trajectory-class membership. To ensure the validity of the logistic regression assumptions, the ISDE was standardized to meet the model's distribution requirements.

All statistical analyses were implemented using R version 4.3.1 and the "lcmm" package version 2.1.0.

-->

## Data sources

We analyzed municipality-level data from 2011 to 2023, focusing on adults aged ≥ 15 years who were diagnosed with T2DM under the public health scheme. This age threshold aligns with the cardiovascular program guidelines, which include patients aged 15 and above. Data on T2DM management were sourced from the Statistical and Health Information Department and accessed through the Chilean Public Health Platform (CHP).[@deis]

We examined two primary outcome indicators:

-   **Diabetic Glycemic Control Coverage (DGCC):** The Proportion of T2DM patients achieving HbA1c \<7% among those diagnosed and monitored by the cardiovascular program.

    **DRSC:** Proportion of T2DM patients receiving annual diabetic retinopathy screening within the diagnosed and monitored T2DM population under the cardiovascular program.

For the DGCC indicator, it was assumed that T2DM patients had at least one HbA1c test to determine if levels were above or below 7%. If the patients were not tested at the end of the 12-month period, a recovery protocol was initiated. After three failed attempts to contact them, the patients were removed from the program and dataset. Thus, untested patients were excluded and not assumed to have HbA1c levels \> 7%. Conversely, patients undergoing annual DRS remain in the cardiovascular healthcare program.

To improve representativeness, we excluded municipalities in the lowest population-size quintile. These smaller municipalities often showed extreme coverage percentages (e.g., 0% or 100%) because of their small T2DM populations, which could have skewed the results.

We examined the associations between class membership coverage and sociodemographic factors including deprivation, urbanicity, and geographical location. Data on potential predictive factors at the municipal level were collected as follows:

-   **Deprivation:** Measured using the Index of Socioeconomic Development (ISED) score developed by the Chilean Observatory of Public Health, which rates deprivation from 0 to 1 (lower scores indicate higher deprivation). The index includes income, education, housing, and sanitation components.[@gattini_comunas_2013]

-   **Urbanicity:** Based on classifications by the National Statistics Institute and following the National Policy of Rural Development Framework.[@bergamini2020] Municipalities were categorized based on population density as predominantly rural (more than 50% of the population residing in low-density areas), mixed (25–49% in low-density areas), and predominantly urban (less than 25% in low-density areas).

-   **Geographical Location:** We used the Environmental Tribunals of Chile classification, which divides the country into northern, central, and southern zones. This classification simplifies the geographical locations of municipalities into three levels.

## Statistical analyses

We performed a two-stage analysis, as follows:

-   LCMM: In the first stage, we employed a Latent Class Mixed Model (LCMM) to identify trajectory classes for DRSC and DGCC from 2011 to 2023. LCMM probabilistically assigns latent classes representing subgroups or unique patterns within a population17. Model complexity was explored from one to seven clusters, hypothesizing that the intercept and slope represent random effects across municipalities. Selection was guided by an eight-step framework (see Table 11), and we assessed model fit through goodness-of-fit measures including AIC, BIC, SABIC, CAIC, AWE, Bayes Factor, and correct model probability (cmP). Diagnostic tests included entropy, smallest class count (SCC), size (SCS), and minimum Average Latent Class Posterior Probability (ACPP), along with the Vuong-Lo-Mendell-Rubin (VL-LRT) test to compare the models. We favored the k-class model over the k-1 class model when the p-value was below 0.05, considering clinical plausibility, model fit, and diagnostic criteria.[@wickrama2021]

-   Class Membership Influences: In second stage, we analyzed whether municipality-level factors (deprivation, urbanicity, and geographical location) influenced the trajectory class membership. The ISED score was standardized after verifying the distributional assumptions for the logistic regression models, and all analyses were conducted using the standardized score. Both adjusted and unadjusted analyses were conducted to validate these associations.

Finally, we conducted a sensitivity analysis covering 2011 to 2019 to assess the potential impact of excluding the pandemic years (2019–2023) on the DRSC and DGCC trajectories. We also assessed data completeness across observations over time, comparing means using Welch’s t-test to ensure balanced data distributions across trajectory classes. All statistical analyses were performed using R version 4.3.1 and the “lcmm” package version 2.1.0.

<!--
The DGCC and DRSC raw trajectories over the twelve-year period are presented in Figure 1 (panel a and b, respectively), showing unclear patterns. Figure 1 depict 300 trajectories for each set, with DGCC appearing to be less heterogeneous than DRSC. Data averaged a 91.08% (10.93 years) of completeness for DGCC and 87.4% (10.5 years) for DRSC over the twelve-year period. Table 1 displays the LCMM results for DGCC and DRSC outcomes. A one-class model using a simple linear model predicts coverage and cluster membership, with intercepts representing baseline coverage and slopes indicating average annual change. For the one-class model (Table 1), the DGC started with a baseline coverage of 47.2% (SE 0.53) and a significant yearly decrease of 1.18% (SE 0.057, *p\<0.001*). For DGCC, the one-class model represents the overall trend, as illustrated in Figure 1c. For DRS, the one-class model exhibited a baseline coverage of 32.3% (SE 1.19) and a slight but significant annual increase of 0.30% (SE 0.14, *p=0.03*), as depicted in Figure 1d.

We selected the one-class DGCC model, which was the most parsimonious and outperformed all others (see Table 1). The two-class model was selected as the preferred solution for the DRSC. Figure 3 depicts this model, where class membership 1 or the "lower and stable DRSC class" (206 municipalities - 68.7%) started at 27.9% (SE 2.51) in 2011 and experienced a slight annual decrease of 0.04% (SE 0.26), whereas class 2 or the "moderate and stable DRSC class" (94 municipalities - 31.3%) began at 42.1% (SE 4.09) and exhibited a slight annual increase of 1.01% (SE 0.57). Wald's test indicated no significant changes over both class 1 (*p=0.863)* and class 2 (*p=0.078*). The average completeness of data for DRSC was 91.16% (10.94 years) for the "lower and stable DRSC class" and 87.16% (10.46 years) for the "moderate and stable DRSC class", respectively, not founded statistically significative differences between their means (*p=0.097*, Welch's t-test).

[Supplementary data](https://github.com/Rolo-Silva/trajectories_diabetes) 1 provide details on trajectory class memberships for each municipality, including the probabilities of belonging to either class 1 or class 2, along with coverage information.

## Sensitivity analysis of estimated trajectories for DGCC and DRSC by excluding COVID-19 pandemic years

In the 2011-2019 period (n=292 municipalities), data averaged a 93.46% (8.41 years) of completeness for DGCC and 89.61% (8.07 years) for DRSC over the nine-year period. The one-class model for DGC displayed an initial coverage of 45.1% (SE 0.53) and revealed a statistically significant yearly decrease of 0.42% (SE 0.07). For the DRS, the one-class model displayed an average coverage of 28.45% (SE 1.16), with an annual increase of 1.59% (SE 0.17). Like the 2011-2023 period, DGCC maintained a single trajectory, whereas DRSC persisted with two trajectories.\
For the DRSC, Class 1 (217 municipalities - 74.32%; lower DRSC) began in 2011 at 23.05% (SE 1.65) and exhibited a gradual increase to 2019 at an annual rate of 1.38% (SE 0.30). In contrast, Class 2 (75 municipalities - 25.68%; higher DRSC) started with a substantial coverage of 43.58% (SE 4.3) in 2011 and experienced a faster increase through 2019 at a rate of 2.16% per annum (SE 0.69). Notably, both trajectories displayed statistically significant slopes (wald's test p\<0.01). The overall completeness of the data for DRSC was 90.17% (8.11 years) for Class 1 and 88.3% (7.95 years) for Class 2. No significant differences were detected between classes in terms of completeness (p=0.524, Welch's t-test).

-->

# Results

## Overview of DGCC and DRSC Latent Class Trajectory Estimation

We analyzed 300 municipalities during the 2011-2023 period and 292 municipalities during the 2011-2019 period. The trajectories of Diabetic Glycemic Control Coverage (DGCC) and Diabetic Retinopathy Screening Coverage (DRSC) over these periods exhibited notable differences. The summary statistics for both outcomes and sociodemographic characteristics are presented in Tables 1 and 2, respectively. As shown in Figure 1 (panels a and b), the raw trajectories of DGCC and DRSC over the 13-year period displayed unclear patterns, with DRSC showing more heterogeneity than DGCC.

Model fit statistics for the Latent Class Mixed Model (LCMM) for both DGCC and DRSC are summarized in Table 3. Baseline DGCC coverage was slightly higher in the 2011-2023 period (46.7%) than in the 2011-2019 period (45.1%). However, both periods exhibited a significant decline in annual coverage: a -1.03% change per year (p \< 0.001) for 2011-2023 and -0.42% per year (p \< 0.001) for 2011-2019. In contrast, DRSC coverage increased from 28.5% in 2011-2019 to 31.3% in 2011-2023. The annual change in DRSC was more pronounced in the shorter period, with an increase of 1.59% per year (p \< 0.001), compared with 0.5% per year (p \< 0.001) in the longer period.

## Latent Class Models

For DGCC, the one-class model was selected as the most parsimonious solution, outperforming all other models for both–2011-2023 and 2011-2019 periods. For the DRSC, the two-class model was chosen as the optimal solution for both periods (Table 3).

Figure 3 presents the LCMM results for DRSC, where two classes are identified:

-   **Class 1** (lower DRSC coverage) showed a stable baseline coverage and minimal annual changes. For the 2011-2023 period, the baseline coverage was 24.8%, with an annual change of 0.42% (p = 0.068). In the 2011-2019 period, Class 1 had a baseline coverage of 23.05%, with an annual change of 1.38% (p \< 0.001), while in the 2020-2023 period, the baseline coverage was 15.8%, with an annual increase of 8.3% (p \< 0.001).

-   **Class 2** (higher DRSC coverage) exhibited higher baseline coverage and notable annual changes. In the 2011-2023 period, the baseline coverage was 43.72%, with an annual change of 0.69% (p = 0.073). In the 2011-2019 period, Class 2 had a baseline coverage of 43.58%, with an annual change of 2.16% (p = 0.0017), while in the 2020-2023 period, the baseline coverage was 47.41%, with an annual increase of 3.82% (p \< 0.016).

These trajectory classes reflect significant differences in both baseline coverage and annual changes over the study period.

## Influential Factors in Diabetic Retinopathy Screening Coverage

We analyzed the effects of various factors on DRSC class membership, and the results are summarized in Table 3. Using a logistic regression model, we explored the associations of deprivation (ISED score), urbanicity, and geographical location with DRSC trajectory-class membership.

### Deprivation

Our analysis revealed significant associations between deprivation and DRSC class membership in both the periods. Higher deprivation (lower ISED scores) was associated with a decreased likelihood of being in a higher DRSC coverage class. The odds ratios (OR) for this association were:

-   **2011-2023 period**: OR = 0.66 (95% CI: 0.50, 0.85)

-   **2011-2019 period**: OR = 0.53 (95% CI: 0.39, 0.72)

-   **2020-2023 period**: OR = 0.67 (95% CI: 0.49, 0.92)

These findings indicate that municipalities with higher deprivation scores (less deprived) are less likely to belong to the higher-coverage DRSC class.

### Urbanicity

Urbanicity has a varied impact on DRSC coverage, with rural areas generally showing higher odds of improved coverage than mixed-urban zones.

-   **2011-2023 period**: Rural areas had an OR of 1.79 (95% CI: 0.99, 3.21), indicating higher odds of being in the higher coverage class compared to mixed urban areas (reference category), although this was not statistically significant. Urban areas had an OR of 0.39 (95% CI: 0.17, 0.84), showing significantly lower odds of belonging to the higher DRSC class compared to mixed urban areas.

-   **2011-2019 period**: Rural areas had an OR of 2.73 (95% CI: 1.41, 5.30), indicating significantly higher odds of being in the higher coverage class compared to mixed zones. Urban areas had an OR of 0.45 (95% CI: 0.18, 1.19), suggesting lower odds of being in the higher coverage class relative to the reference group, although this was not significant.

-   **2020-2023 period**: Rural areas had an OR of 2.26 (95% CI: 1.08, 4.74), whereas urban areas had an OR of 0.48 (95% CI: 0.17, 1.37) relative to mixed areas.

Overall, rural municipalities tended to have higher DRSC coverage, whereas urban municipalities showed more variable results.

### Geographical Location

Geographical location significantly influenced DRSC class membership, with the central region serving as the reference category because of its representative demographic profile and generally better access to healthcare service provision.[@núñez2021]

-   **2011-2023 period**: The southern region had an OR of 1.85 (95% CI: 1.10, 3.08), compared to the central region (reference category), while the northern region had a relatively neutral impact which had an OR of 1.02 (95% CI: 0.42, 2.51).

-   **2011-2019 period**: The southern region had a higher odds ratio of 1.92 (95% CI: 1.11, 3.34), suggesting a strong positive association with higher DRSC coverage, while the northern region had a relatively neutral impact which had an OR of 0.59 (95% CI: 0.19, 1.83))

-   **2020-2023 period**: The southern region had a higher odds ratio of 1.32 (95% CI: 0.72, 2.43), suggesting positive association with higher DRSC coverage but not significative, while the northern region had a relatively neutral impact.

These findings indicate that municipalities in the southern region are more likely to achieve higher DRSC coverage than other regions.

<!--
## Unadjusted Associations

### Deprivation

In both the 2011-2023 and 2011-2019 periods, the unadjusted analysis revealed a significant association between the standardized index and class membership. Higher values of the standardized index were consistently associated with a decreased likelihood of belonging to the higher coverage class (OR = 0.665, 95% CI \[0.506, 0.864\], p = 0.00276 for 2011-2023; OR = 0.665, 95% CI \[0.506, 0.864\], p = 0.00276 for 2011-2019 for 2011-2019).

### Urbanicity

Similarly, classification type showed significant associations in unadjusted models for both periods. Municipalities classified as Rural were consistently more likely to belong to the higher coverage class compared to those classified as Mixed (OR = 2.10, 95% CI \[1.16, 3.88\], p = 0.016 for 2011-2023; OR = 2.15, 95% CI \[1.27, 3.72\], p = 0.005 for 2011-2019). The Urban classification exhibited a negative association with higher coverage membership, although non-significant (OR = 0.47, 95% CI \[0.21, 1.01\], p = 0.058 for 2011-2023; OR = 0.58, 95% CI \[0.32, 1.05\], p = 0.071 for 2011-2019)

### Geographical location

This variable also demonstrated significant associations in the unadjusted analysis for both periods. While being in the North zone showed a positive but non-significant association with higher coverage membership compared to the central zone, individuals in the South zone exhibited a positive association, though not statistically significant (OR = 1.51, 95% CI \[0.904, 2.53\], p = 0.116 for 2011-2023; OR = 1.45, 95% CI \[0.83, 2.51\], p = 0.189 for 2011-2019).

## Multivariate Associations

### Incorporating Multiple Variables

When adjusting for multiple variables, including urbanicity and geographical location, the association found between the socio-ecnonomic development and class membership in the unudjusted model lost significance in both periods (OR = 1.02, 95% CI \[0.706, 1.47\], p = 0.903 for 2011-2023; OR = 1.02, 95% CI \[0.706, 1.47\], p = 0.903 for 2011-2019). This suggests that the predictive power of the deprivation is influenced by the inclusion of other factors.

### Persisting Effects of Classification and Zona

Despite the loss of significance for the standardized index in the adjusted models, the Rural classification retained its significance in predicting higher coverage membership for both periods (OR = 2.18, 95% CI \[1.16, 4.20\], p = 0.0168 for 2011-2023; OR = 2.24, 95% CI \[1.27, 3.99\], p = 0.005 for 2011-2019), indicating its robust influence over time. Additionally, trends towards significance for the Urbana classification and the positive association of the Sur zone with higher coverage membership persisted, although not statistically significant.

### Persisting Effects of Urbanicity and Zona

Despite the loss of significance for the standardized index in the adjusted model, the Rural classification retained its significance, though with a slight attenuation in effect size (OR = 2.18, 95% CI \[1.16, 4.20\], p = 0.0168). Additionally, the classification type continued to show trends towards significance, while the zona variable did not maintain significant associations in the adjusted models.

Furthermore, the additional analysis covering the 2011-2019 period reaffirmed the findings from the extended 2011-2023 period. In the unadjusted analysis, significant associations were observed for predictors of higher coverage membership. Individuals classified as Rural were more likely to belong to the higher coverage class compared to those classified as Mixta (OR = 2.15, 95% CI \[1.27, 3.72\], p = 0.005). The Urbana classification showed a negative association with higher coverage membership, albeit non-significant (OR = 0.58, 95% CI \[0.32, 1.05\], p = 0.071). Regarding the zona variable, individuals in the Sur zone exhibited a positive association with higher coverage membership, although not statistically significant (OR = 1.45, 95% CI \[0.83, 2.51\], p = 0.189).

When adjusting for multiple variables, including classification and zona, the associations observed in the unadjusted analysis persisted. The Rural classification remained significant (OR = 2.24, 95% CI \[1.27, 3.99\], p = 0.005), indicating its robust influence on higher coverage membership. Similarly, the trends towards significance for the Urbana classification and the positive association of the Sur zone with higher coverage membership persisted, although not statistically significant.

-->

# Discussions

## Summary of Key Findings

We used unsupervised machine learning to identify different subpopulation trajectory patterns for DGCC and DRSC in Chilean municipalities over 12 years. For both the 2011-2023 and 2011-2019 periods, the trajectory patterns for DRSC revealed two distinct patterns. The first depicted municipalities with stable baseline coverage levels, hovering at 27.85% (SE: 2.51) in the 2011-2023 period and slightly lower at 23.05% (SE=1.65%) in the 2011-2019 period. The second trajectory consistently showed higher baseline coverage levels, at 42.14% in the 2011-2023 period and 43.58% in the 2011-2019 period, with significant annual increases in coverage ranging from 1-2%.

Higher deprivation levels (less deprived) were consistently associated with lower DRSC. Rural areas generally showed better coverage than mixed urban environments, while urban areas displayed mixed results. Geographically, municipalities in the southern region consistently exhibited higher DRSC odds ratios compared to the central region. In the adjusted analysis, the influence of urbanicity emerged as the most salient predictor, with rural areas consistently demonstrating higher odds of better coverage.

## What is already known on this topic

Using the same dataset, we have identified significant variation in linear trends in DRSC between Chilean municipalities over the past decade. On the one hand, while deprivation is a predictor for poor uptake, the use of telemedicine has been shown to improve access to diabetic eye care in underserved communities, which often include deprived areas.[@rawlingschid2024] Deprivation affects not only the ability to afford healthcare but also the availability and utilization of services, as it is often associated with geographic isolation and limited healthcare infrastructure[@lewandohundt2012] **(Hundt et al., 2011; Lovett et al., 2004)**.

In our study, we found that areas characterized by higher levels of deprivation surprisingly exhibited better DRSC. This unexpected finding challenges conventional assumptions and suggests that despite facing socioeconomic challenges, these areas may have prioritized healthcare resources towards DRS services. Possible explanations could include targeted healthcare interventions, focused on the importance of regular screening in deprived communities. Certainly, the use of telemedicine and the adoption of more than 150 primary eye-care units across the country may play a significant role. Deprivation may still be relevant in understanding DRS coverage, but its independent effect may be better explained by other factors included in the model as urbanicity. In fact, deprivation lost its significance after adjusting for urbanicity. While some studies indicate that rural areas may have lower screening coverage compared to urban areas,[@tóth2019] other initiatives have demonstrated the feasibility and effectiveness of targeted screening programs in rural settings, [@hall2022; @khou2021] suggesting that targeted interventions can overcome barriers to screening coverage.[@hall2022] In other words, while socioeconomic status alone may be associated with higher coverage membership, its effect is mediated or influenced by factors such as rural/urban level.

This technique has increased interest in health services for obesity,[@viner2018] food addiction,[@aloi2024] loneliness in adolescence,[@verboon2022] risk patterns of myopia[@li2023], or treatment of substance addiction disorders.[@bórquez2024]

The trajectories of DGCC and DRSC were observed to differ from one another. We found DGCC to present a single trajectory encompassing all 301 municipalities, whereas DRSC was characterized by two probable trajectories without significant slopes between 2011 and 2023. While the overall trend for DGCC indicated a decreasing adequate glycemic control over time, the DRSC trajectories showed an increasing trend, albeit still far from an ideal target - 80% in coverage- to tackle sight loss to long term as been reported.[@Scanlon_2017] It should be noted that the COVID-19 pandemic may have had a disruptive impact on healthcare services, including diabetes monitoring and screening,[@ahmed2021] and could have influenced trend analyses. Excluding the pandemic period yielded consistent results but with significant changes in trajectories over time for the DRSC. It is important to exercise caution when interpreting longitudinal trends that may have been influenced by pandemic data, as demonstrated by our sensitivity analyses.

In the unadjusted analysis, we observed a significant association between the standardized index and class membership. This finding underscores the importance of considering the socioeconomic status, as represented by the standardized index, in predicting membership in the higher coverage class. The standardized index serves as a proxy for various socioeconomic factors that may influence an individual's access to healthcare services, such as income level, education, and employment status.

However, when we adjusted for multiple variables, including classification type and zona, the significance of the association between the standardized index and class membership disappeared. This suggests that the predictive power of the standardized index is confounded by the inclusion of other factors. In other words, while socioeconomic status alone may be associated with higher coverage membership, its effect is mediated or influenced by factors such as rural/urban classification and geographical location (zona).

For example, individuals classified as Rural may have lower socioeconomic status on average compared to those classified as Urban, which could contribute to the observed association between the standardized index and class membership. However, once we account for the Rural classification in the multivariate analysis, the direct effect of the standardized index becomes less significant. This indicates that much of the variation in higher coverage membership explained by socioeconomic status is already captured by the classification variable.

Therefore, while the standardized index provides valuable insight into the socioeconomic determinants of higher coverage membership, its significance in predicting membership diminishes when other relevant factors are taken into consideration. This highlights the need for comprehensive multivariate analyses to better understand the complex interplay between socioeconomic status and other predictors of healthcare coverage.

Interestingly, higher levels of deprivation did not lead to less favorable trajectories for municipalities. Instead, they were associated with an increased likelihood of a high DRSC coverage. This association is particularly pronounced between the period 2011-2019. Although it remained highly significant in the 2011-2023 timeframe, the OR exhibited a smoother association, robustly supporting our findings. On the other hand, unadjusted association with urbanicity and their three categories showed that the Urban population is the less favourable in achieving the highuer coverage trajectory, meanwhile Rural population the most favourable, though non statistically significant. It seems urbanisation could explain better the association with DRSC trajectories than deprivation as this lost its significant effect when we adjusted by urbanisation level. Therefore, understanding the factors that influence DRSC is crucial. Future research should explore these practical implications. Potentially, factors linked to targeted health policies addressing diabetes within Chile's 2005 health reform[@bastias2008] and financing primary healthcare based on factors such as deprivation through capitation[@cuadrado2022] could provide insights and explain, at least in part, our findings.

To date, there is no evidence of the use of LCMM techniques to investigate geographical differences in the DRSC in Chile or other regions. Similar evidence has explored obesity prevalence in childhood across local authorities in the UK, highlighting its significance in studying geographical data to inform policies.[@viner2018] Despite previous discussions, no studies at the municipal level have used mixed-effect models or LCMM techniques to identify unique trends in the DRSC. LCMM methods enable the identification of municipalities that demonstrate exceptional progress in longitudinal trends for sight-loss-related diabetes prevention, enhance the understanding of geographical inequalities, and inform targeted interventions to improve screening coverage.

Utilizing PSCV data from more than a decade, we successfully identified distinct subpopulations of municipalities that exhibited varying trajectories in the DRSC. Most municipalities displayed a high probability of belonging to their designated classes, suggesting a strong alignment between these assigned classes and DRSC from 2011 to 2023. However, some municipalities exhibited notably lower probabilities of being assigned to the DRSC classes, indicating potential inaccuracies. Although our analyses were conducted at the ecological (municipality-level) rather than at the individual level (T2DM patient), this approach was considered appropriate as a key determinant of healthcare access for DRSC often operating at the primary healthcare level. Our primary objective was to identify the potential subpopulations among municipalities. Nevertheless, we could not establish causality for the relationships identified between socioeconomic deprivation and municipality trajectory classes.

While the trajectory classes we identified are plausible, LCMM is an exploratory, data-driven technique, and chance relationships in our data may have influenced our findings regarding trajectory groups. However, we were cautious in identifying the most likely latent class groups and the trajectories we identified were reasonable. The southern regions exhibited a notable trend, with a higher likelihood of falling into the upper DRSC category, suggesting distinct regional patterns potentially influenced by socioeconomic conditions or healthcare accessibility in these areas. Further investigation into the impact of regional policies on DRSC access could be beneficial, encompassing factors such as the availability of eye care specialists, the implementation of mobile clinics for underserved populations, and targeted initiatives to enhance access in less advantaged regions. These area-specific efforts could play a crucial role in broadening screening coverage and mitigating regional healthcare inequalities.

Our data on municipality deprivation took place in 2013, but we hypothesized that it is possible that the overall deprivation levels remained relatively stable over the thirteen-year (and nine-year) study period. Nonetheless, trajectory class membership may have been influenced by changes in deprivation levels during the study period.

# Conclusions

The existence of DGCC in Chilean municipalities appears to follow a single trajectory, whereas the DRSC presents two distinct, gradually increasing trajectories after testing the pandemic period. Higher deprivation levels (lower ISED score) were associated with an increased likelihood of higher DRSC trajectory. It is worthwhile to examine municipalities that demonstrate a higher DRSC trajectory (see [supplementary data](https://github.com/Rolo-Silva/trajectories_diabetes)) in greater detail to identify any shared local policies or practices that may contribute to their positive deviation from national trends in DRSC increase. Future studies should investigate the factors that influence DRSC coverage classes over time, and consider the implications of enhancing health systems in relation to DGCC/DRSC trajectories.

**Conflict of interest**

Authors declared they have no conflicts of interest.

**Funding**

None.

**Acknowledgements**

**Author contribution.**

```{r}
#| label: fig-1
#| fig-cap: "Raw data coverage trajectories 2011-2023"
#| fig-subcap: 
#| - "Diabetic Glycemic Control Coverage"
#| - "Diabetic Retinopathy Screening Coverage"
#| layout-ncol: 2
#| fig-align: center

# Plots for raw trajectories for DRSC and DGCC --------------------

coverage1 <- read.csv("coverage_2011_2023_noq1.csv", sep=",") #read coverage file
#coverage2 <- read.csv("coverage_2011_2019_noq1.csv", sep=",")


ggplot(coverage1, aes(x=ano2, y=dm_coverage)) +
  geom_point(size = .2) + 
  geom_line(size = .2, aes(group = comuna2)) +
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM individuals with HbA1C < 7%")+
  #theme_ipsum()+
 theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  theme(legend.position = "none") +
  #geom_smooth(se = T, method = "loess", aes(group = class))+
  scale_color_lancet()


ggplot(coverage1, aes(x=ano2, y=drs_coverage)) +
  geom_point(size = .2) + 
  geom_line(size = .2, aes(group = comuna2)) +
scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  labs(colour= "class magnitude")+
  theme(legend.position = "right") +
  #geom_smooth(se = T, method = "loess", aes(group = class))+
 # scale_color_lancet(labels = c("82.27%", "17.73%")) 
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  scale_color_lancet()




```

```{r}
#| label: fig-2
#| warning: false
#| fig-cap: "Baseline one class trend data coverage trajectories 2011-2023"
#| fig-subcap: 
#| - "Diabetic Glycemic Control Coverage"
#| - "Diabetic Retinopathy Screening Coverage"
#| layout-ncol: 2
#| fig-align: center

dm1 <- read.csv("dm_2011_2023_noq1.csv", sep=",") 

ggplot(dm1, aes(x=ano2, y=dm_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM individuals with HbA1C < 7%")+
  #theme_ipsum()+
  theme(legend.position = "none") +
  geom_smooth(se = T, method = "loess", aes(group = class))+
   theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  scale_color_lancet()

dr1 <- read.csv("drs_2011_2023_noq1.csv", sep=",") 

ggplot(dr1, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  scale_color_lancet()
  

```

```{r}
#| label: fig-3
#| warning: false
#| fig-cap: "Latent class trajectory mixture models solutions for diabetic retinopathy screening coveragein three differente periods; 2011-2023; 2011-2019; and 2020-2023"
#| fig-subcap: 
#| - "2011-2023"
#| - "2011-2019"
#| - "2020-2023"
#| layout-ncol: 1
#| fig-align: center

dr2 <- read.csv("drs2_2011_2023_noq1.csv", sep=",") 

ggplot(dr2, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        #legend.box.background = element_rect(colour = "red"),
        legend.key = element_rect(fill = "white", colour = "white"))+
  #scale_color_lancet()+
  scale_color_lancet(labels = c("202 trajectories (67.3%)", "98 trajectories (32.7%)")) 


dr3 <- read.csv("drs2_2011_2019_noq1.csv", sep=",") 

ggplot(dr3, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        #legend.box.background = element_rect(colour = "red"),
        legend.key = element_rect(fill = "white", colour = "white"))+
  #scale_color_lancet()+
  scale_color_lancet(labels = c("217 trajectories (74.3%)", "75 trajectories (25.7%)")) 


dr4 <- read.csv("drs2_2020_2023_noq1.csv", sep=",") 

ggplot(dr4, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        #legend.box.background = element_rect(colour = "red"),
        legend.key = element_rect(fill = "white", colour = "white"))+
  #scale_color_lancet()+
  scale_color_lancet(labels = c("245 trajectories (81.7%)", "55 trajectories (18.3%)")) 

```

```{r}
#| label: tbl-1
#| tbl-cap: "Numeric variables summary of diabetes coverage across two periods"
#| tbl-subcap: 
#| - "Descriptive statistics by municipalities (Municipality data average)"
#| - "Descriptive statistics of all raw data (Obsevation-based data)"
#| fig-align: center

#The mean of the means


# Read data
d1 <- read.csv("data_reglog.csv", sep=",")
d2 <- read.csv("data_reglog2.csv", sep=",")


# Summarize data
data_summary1 <- modelsummary::datasummary_skim(d1[1:300,] %>% 
                                                  select("mean_dm_coverage", 
                                                         "mean_drs_coverage", 
                                                         "index_standardized") %>% 
                                                  mutate_if(is.character, as.factor),
                                                type="numeric",
                                                histogram=T,
                                                fmt = "%.4f",
                                                output="gt") %>% 
  as.data.frame() 

data_summary2 <- modelsummary::datasummary_skim(d2[1:292,] %>% 
                                                  select("mean_dm_coverage", 
                                                         "mean_drs_coverage", 
                                                         "index_standardized") %>% 
                                                  mutate_if(is.character, as.factor),
                                                type="numeric",
                                                histogram=T,
                                                fmt = "%.4f",
                                                output="gt") %>% 
  as.data.frame()

data_summary3 <- rbind(data_summary2, data_summary1)

#data_summary3 <- data_summary3 %>%
 # dplyr::mutate(Observations = dplyr::case_when(
  #  dplyr::row_number() <= 3 ~ 292,
   # TRUE ~ 300
 # ))


# Check column types and convert if necessary

data_summary3 <- data_summary3 %>% 
  # Convert columns 4 to 8 to numeric
  mutate(across(4:8, ~ as.numeric(.))) %>%  
  # Conditionally apply rounding and multiplication by 100 to columns 4 to 8, excluding rows 3 and 6
  mutate(across(4:8, ~ case_when(
    row_number() %in% c(3, 6) ~ .,       # Keep original values in rows 3 and 6
    TRUE ~ round(. * 100, 2)             # Round other rows after multiplying by 100
  ))) %>%
  # Add the Metric column
  dplyr::mutate(Metric = rep(c("DGCC", "DRSC", "ISDE"), 2)) %>%
  # Apply a final rounding to two decimals for all rows in columns 4 to 8
  mutate(across(4:8, ~ round(., 2)))


 
#colnames(data_summary3)[9] <- "Histogram"

# Print table using gt
data_summary3 %>% 
  dplyr::select(Metric, -1, 2:9) %>% 
  gt() %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>% 
  tab_row_group(
    group = "2011-2023",
    rows = 4:6
  ) %>%
  tab_row_group(
    group = "2011-2019",
    rows = 1:3
  )%>% 
 tab_source_note(source_note = md(" **Abbreviations:** DGCC - Diabetic glycemic control coverage; DRSC - Diabetic retinopathy screening coverage, ; ISDE - Index of Socioeconomic Development"))

# Summarize data for dr2
data_summary4 <- modelsummary::datasummary_skim(dr2 %>% 
                                                  select("dm_coverage", 
                                                         "drs_coverage"),
                                                type="numeric",
                                                histogram=T,
                                                fmt = "%.4f",
                                                output="gt") %>% 
  as.data.frame() 


# Summarize data for dr2
data_summary5 <- modelsummary::datasummary_skim(dr2 %>% 
                                                  select("dm_coverage", 
                                                         "drs_coverage"),
                                                type="numeric",
                                                histogram=T,
                                                fmt = "%.4f",
                                                output="gt") %>% 
  as.data.frame() 



# Combine summaries
data_summary6 <- rbind(data_summary5, data_summary4)

# Print the intermediate result to debug column numbers
#print(data_summary6)

# Check column types and convert if necessary


data_summary6 <- data_summary6 %>% 
  mutate(across(4:8, ~ as.numeric(.))) %>% # Convert to numeric
  mutate(across(4:8, ~ round(. * 100, 2)))  %>% 
  dplyr::mutate(Metric = rep(c("DGCC", "DRSC"), 2))# Apply rounding 
 
#colnames(data_summary6)[9] <- "Histogram"


# Print table using gt
data_summary6 %>% 
  dplyr::select(Metric, -1, 2:9) %>% 
  gt() %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>% 
  tab_row_group(
    group = "2011-2023",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "2011-2019",
    rows = 1:2
  )%>% 
  tab_source_note(source_note = md(" **Abbreviations:** DGCC - Diabetic glycemic control coverage; DRSC - Diabetic retinopathy screening coverage, ; ISDE - Index of Socioeconomic Development"))

```

```{r}
#| label: tbl-2
#| tbl-cap: "Summary statistics of diabetes coverage and sociodemographics in Chilean municipalities across two periods"
#| fig-align: center


zone_summary <- d1[1:300, ] %>%
  group_by(zona) %>%
  dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(d1[1:300, ]) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic =zona)



classification_summary <- d1[1:300, ] %>%
  group_by(classification) %>%
   dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(d1[1:300, ]) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic = classification)



sociodemographics1 <- rbind(zone_summary, classification_summary)


zone_summary2 <- d2[1:292, ] %>%
  group_by(zona) %>%
    dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(d2[1:292, ]) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic =zona)


classification_summary2 <- d2[1:292, ] %>%
  group_by(classification) %>%
    dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(d2[1:292, ]) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic = classification)


sociodemographics2 <- rbind(zone_summary2, classification_summary2)

sociodemographics <- rbind(sociodemographics2, sociodemographics1)

sociodemographics <- sociodemographics %>%  
  select(Demographic, N_comuna, percent_comuna, 
         longitudinal_mean_dm_coverage, longitudinal_sd_dm_coverage, 
         longitudinal_mean_drs_coverage, longitudinal_sd_drs_coverage,
         mean_index_standardized, sd_index_standardized) %>% 
  mutate(across(4:7, ~ . * 100)) %>% 
  mutate_if(is.numeric, round,2) %>% 
  mutate(Demographic = case_when(
    Demographic == "centro" ~ "Center",
    Demographic == "norte" ~ "North",
    Demographic == "sur" ~ "South",
    Demographic == "Mixta" ~ "Mixed",
    Demographic == "Rural" ~ "Rural",
    Demographic == "Urbana" ~ "Urban",
    TRUE ~ Demographic
  ))
 

sociodemographics %>% 
  gt::gt() %>%
  #tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white") %>% 
  tab_row_group(group = "Zone (2011-2023)",
rows = c(7:9)
  ) %>%
  tab_row_group(group = "Urbanicity (2011-2023)",
rows = c(10:12)
  ) %>%
  tab_row_group(
    group = "Zone (2011-2019)",
rows = c(1:3)
  )  %>%
  tab_row_group(
    group = "Urbanicity (2011-2019)",
rows = c(4:6)
  ) %>% 
  cols_label(N_comuna = "N",
             percent_comuna = "%",
             longitudinal_mean_drs_coverage = "Mean",
             longitudinal_sd_drs_coverage = "SE",
             longitudinal_mean_dm_coverage = "Mean",
             longitudinal_sd_dm_coverage = "SE",
             mean_index_standardized = "Mean",
             sd_index_standardized = "SE") %>%
  tab_spanner(
    label = "ISDE",
    columns = c("mean_index_standardized", "sd_index_standardized")
  ) %>% 
  tab_spanner(
    label = "DGCC",
    columns = c("longitudinal_mean_dm_coverage", "longitudinal_sd_dm_coverage")
  ) %>% 
  tab_spanner(
    label = "DRSC",
    columns = c("longitudinal_mean_drs_coverage", "longitudinal_sd_drs_coverage")
  ) %>% 
  tab_spanner(
    label = "Frequency",
    columns = c("N_comuna", "percent_comuna")
  ) %>% 
   tab_source_note(source_note = md(" **Abbreviations:** DGCC - Diabetic glycemic control coverage; DRSC - Diabetic retinopathy screening coverage; ISDE - Index of Socioeconomic Development; N - number of occurrences or observations within each category."))




```

```{r}
#| label: tbl-3
#| tbl-cap: "Latent class trajectory mixture models for diabetic glycemic control and diabetic retinopathy screening coverage across 300 (2011-2023) and 292 (2011-2019) municipalities"
#| fig-align: center

#DGCC 2011-2023
mfc1 <- read.csv("model_fit_criteria_dm_2011_2023_noq1.csv", sep=",") 

mfc1 <- mfc1 %>% 
  dplyr::rename("LCMM class" = G, 
                "Maximum log-likelihood"= loglik,
                "Nº of parameters"=npm,
                "Sample size"= sample_size)

dc1 <- read.csv("diagnostic_criteria_dm_2011_2023_noq1.csv", sep=",") 

dc1 <- dc1 %>% 
  dplyr::rename("LCMM class" = G, 
                "Smallest class count (n)"= smallest_class_count,
                "Smallest class size (%)"= smallest_class_size_perc)

model_criteria <- left_join(mfc1, dc1)

#DGCC 2011-2019

mfc2 <- read.csv("model_fit_criteria_dm_2011_2019_noq1.csv", sep=",") 

mfc2 <- mfc2 %>% 
  dplyr::rename("LCMM class" = G, 
                "Maximum log-likelihood"= loglik,
                "Nº of parameters"=npm,
                "Sample size"= sample_size)

dc2 <- read.csv("diagnostic_criteria_dm_2011_2019_noq1.csv", sep=",") 

dc2 <- dc2 %>% 
  dplyr::rename("LCMM class" = G, 
                "Smallest class count (n)"= smallest_class_count,
                "Smallest class size (%)"= smallest_class_size_perc)

model_criteria2 <- left_join(mfc2, dc2)

#DGCC 2020-2023

mfc3 <- read.csv("model_fit_criteria_dm_2020_2023_noq1.csv", sep=",") 

mfc3 <- mfc3 %>% 
  dplyr::rename("LCMM class" = G, 
                "Maximum log-likelihood"= loglik,
                "Nº of parameters"=npm,
                "Sample size"= sample_size)

dc3 <- read.csv("diagnostic_criteria_dm_2020_2023_noq1.csv", sep=",") 

dc3 <- dc3 %>% 
  dplyr::rename("LCMM class" = G, 
                "Smallest class count (n)"= smallest_class_count,
                "Smallest class size (%)"= smallest_class_size_perc)

model_criteria3 <- left_join(mfc3, dc3)

table_dgcc <- rbind(model_criteria, model_criteria2)
table_dgcc <- rbind(table_dgcc, model_criteria3)



#DRSC 2011-2023
mfc4 <- read.csv("model_fit_criteria_drs_2011_2023_noq1.csv", sep=",") 

mfc4 <- mfc4 %>% 
  dplyr::rename("LCMM class" = G, 
                "Maximum log-likelihood"= loglik,
                "Nº of parameters"=npm,
                "Sample size"= sample_size)

dc4 <- read.csv("diagnostic_criteria_drs_2011_2023_noq1.csv", sep=",") 

dc4 <- dc4 %>% 
  dplyr::rename("LCMM class" = G, 
                "Smallest class count (n)"= smallest_class_count,
                "Smallest class size (%)"= smallest_class_size_perc)

model_criteria4 <- left_join(mfc4, dc4)

#DRSC 2011-2019

mfc5 <- read.csv("model_fit_criteria_drs_2011_2019_noq1.csv", sep=",") 

mfc5 <- mfc5 %>% 
  dplyr::rename("LCMM class" = G, 
                "Maximum log-likelihood"= loglik,
                "Nº of parameters"=npm,
                "Sample size"= sample_size)

dc5 <- read.csv("diagnostic_criteria_drs_2011_2019_noq1.csv", sep=",") 

dc5 <- dc5 %>% 
  dplyr::rename("LCMM class" = G, 
                "Smallest class count (n)"= smallest_class_count,
                "Smallest class size (%)"= smallest_class_size_perc)

model_criteria5 <- left_join(mfc5, dc5)


#DRSC 2020-2023

mfc6 <- read.csv("model_fit_criteria_drs_2020_2023_noq1.csv", sep=",") 

mfc6 <- mfc6 %>% 
  dplyr::rename("LCMM class" = G, 
                "Maximum log-likelihood"= loglik,
                "Nº of parameters"=npm,
                "Sample size"= sample_size)

dc6 <- read.csv("diagnostic_criteria_drs_2020_2023_noq1.csv", sep=",") 

dc6 <- dc6 %>% 
  dplyr::rename("LCMM class" = G, 
                "Smallest class count (n)"= smallest_class_count,
                "Smallest class size (%)"= smallest_class_size_perc)

model_criteria6 <- left_join(mfc6, dc6)

table_drsc <- rbind(model_criteria4, model_criteria5)

table_drsc <- rbind(table_drsc, model_criteria6)


class_solutions <- rbind(table_dgcc, table_drsc)

class_solutions <- class_solutions %>% 
       dplyr::mutate(across(c(2,4:6,8:13,16:18), round, 2)) %>% 
      dplyr::rename("SCC (n)"= "Smallest class count (n)",
                    "SCS (%)"= "Smallest class size (%)",
                    "SS" = "Sample size",
                    "MLL"= "Maximum log-likelihood",
                    "Parameters (n)" = "Nº of parameters",
                    "Lowest APPA"= Lowest.APPA,
                    "Highest MMV"=Highest.MMV,
                    "Lowest OCC"= Lowest.OCC,
                    "Relative Entropy" = entropy,
                    ) %>% 
  select(1,3,2,4:6,-7,8:19)
  

#knitr::kable(mfc1, row.names = F, digits= 2)

#knitr::kable(dc1, row.names = F, digits= 2)

gt::gt(class_solutions) %>%
#tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
tab_options(
table.font.size = 10,
data_row.padding = px(1),
table.border.top.color = "black",
heading.border.bottom.color = "black",
row_group.border.top.color = "black",
row_group.border.bottom.color = "white",
table.border.bottom.color = "white",
column_labels.border.top.color = "black",
column_labels.border.bottom.color = "black",
table_body.border.bottom.color = "black",
table_body.hlines.color = "white")%>%
  tab_row_group(
    group = "Diabetic retinopathy screening coverage 2020-2023 
(n= 300 municipalities)",
    rows = 31:36 
) %>% 
  tab_row_group(
    group = "Diabetic retinopathy screening coverage 2011-2019 
(n= 292 municipalities)",
    rows =25:30
  ) %>% 
  tab_row_group(
    group = "Diabetic retinopathy screening coverage 2011-2023 
(n= 300 municipalities)",
    rows =19:24
  ) %>% 
   tab_row_group(
    group = "Diabetic glycemic control coverage 2020-2023 
(n= 300 municipalities)",
    rows = 13:18
  ) %>% 
    tab_row_group(
    group = "Diabetic glycemic control coverage 2011-2019 
(n= 292 municipalities)",
    rows = 7:12
 ) %>% 
   tab_row_group(
    group = "Diabetic glycemic control coverage 2011-2023 
(n= 300 municipalities)",
    rows = 1:6
 ) %>%
   tab_source_note(source_note = md(" **Abbreviations:** LCMM - Latent Class Mixture Model; MLL – Maximum Log-likelihood; AIC- Akaike Information Criteria; BIC - Bayesian Information Criteria; SABIC - Sample-Adjusted Bayesian Information Criteria; CAIC -  Consistent Akaike Information Criterion; AWE- Approximate Weight of Evidence Criterion; SIC - Schwarz Information Criterion; BF - Bayes Factor; cmP- correct model probability; SCC - Smallest Class Count; SCS - Smallest Class Size; APPA: Average Class Posterior Probability; MMV: Mistmatch Value; OCC: Odds od Correct Classification; VLMRLRT: Vuong–Lo–Mendell–Rubin Likelihood Radio Test"))
 

```

```{r}
#| label: tbl-4
#| tbl-cap: "Diabetic glycemic control coverage trajectory class characteristics 2011-2019" 
#| fig-align: center

df_dm_2011_2023 <- read.csv("df_dm_2011_2023.csv", sep=",") 

df_dm_2011_2023 <- df_dm_2011_2023 %>% 
  select(-1) 

# Round numeric p-values
df_dm_2011_2023 <- df_dm_2011_2023 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_dm_2011_2023 <- df_dm_2011_2023 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_dm_2011_2023[is.na(df_dm_2011_2023)] <- ""

# Rename columns for display
colnames(df_dm_2011_2023) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_dm_2011_2023) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))



```

```{r}
#| label: tbl-5
#| tbl-cap: "Diabetic glycemic control coverage trajectory class characteristics 2011-2023" 
#| fig-align: center


df_dm_2011_2019 <- read.csv("df_dm_2011_2019.csv", sep=",") 

df_dm_2011_2019 <- df_dm_2011_2019 %>% 
  select(-1) 

# Round numeric p-values
df_dm_2011_2019 <- df_dm_2011_2019 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_dm_2011_2019 <- df_dm_2011_2019 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_dm_2011_2019[is.na(df_dm_2011_2019)] <- ""

# Rename columns for display
colnames(df_dm_2011_2019) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_dm_2011_2019) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))







```

```{r}
#| label: tbl-6
#| tbl-cap: "Diabetic retinopathy screening coverage trajectory class characteristics 2011-2019" 
#| fig-align: center


df_drs_2011_2019 <- read.csv("df_drs_2011_2019.csv", sep=",") 

df_drs_2011_2019 <- df_drs_2011_2019 %>% 
  select(-1) 

# Round numeric p-values
df_drs_2011_2019 <- df_drs_2011_2019 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_drs_2011_2019 <- df_drs_2011_2019 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_drs_2011_2019[is.na(df_drs_2011_2019)] <- ""

# Rename columns for display
colnames(df_drs_2011_2019) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_drs_2011_2019) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))


```

```{r}
#| label: tbl-7
#| tbl-cap: "Diabetic retinopathy screening coverage trajectory class characteristics 2011-2023" 
#| fig-align: center



df_drs_2011_2023 <- read.csv("df_drs_2011_2023.csv", sep=",") 

df_drs_2011_2023 <- df_drs_2011_2023 %>% 
  select(-1) 

# Round numeric p-values
df_drs_2011_2023 <- df_drs_2011_2023 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_drs_2011_2023 <- df_drs_2011_2023 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_drs_2011_2023[is.na(df_drs_2011_2023)] <- ""

# Rename columns for display
colnames(df_drs_2011_2023) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_drs_2011_2023) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))


```

```{r}
#| label: tbl-8
#| tbl-cap: "Diabetic retinopathy screening coverage trajectory class characteristics 2020-2023" 
#| fig-align: center



df_drs_2020_2023 <- read.csv("df_drs_2020_2023.csv", sep=",") 

df_drs_2020_2023 <- df_drs_2020_2023 %>% 
  select(-1) 

# Round numeric p-values
df_drs_2020_2023 <- df_drs_2020_2023 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_drs_2020_2023 <- df_drs_2020_2023 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_drs_2020_2023[is.na(df_drs_2020_2023)] <- ""

# Rename columns for display
colnames(df_drs_2020_2023) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_drs_2020_2023) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))


```

```{r}
#| label: tbl-9
#| tbl-cap: "Logistic regression 2011-2019"
#| fig-align: center
#| results: "asis"



# Model 5: Single predictor (index_standardized)
model5 <- glm(class_membership ~ index_standardized, family = "binomial", data = d2)
# Model 5a: Single predictor (classification)
model5a <- glm(class_membership ~ classification, family = "binomial", data = d2)
# Model 5b: Single predictor (zona)
model5b <- glm(class_membership ~ zona, family = "binomial", data = d2)
# Model 5c: All predictors
model5c <- glm(class_membership ~ index_standardized + classification + zona, family = "binomial", data = d2)
# Model 6: Two predictors (index_standardized + classification)
model6 <- glm(class_membership ~ index_standardized + classification, family = "binomial", data = d2)
# Model 6a: Two predictors (index_standardized + zona)
model6a <- glm(class_membership ~ index_standardized + zona, family = "binomial", data = d2)
# Model 6b: Two predictors (classification + zona)
model6b <- glm(class_membership ~ classification + zona, family = "binomial", data = d2)



models2 <- list("M1" = model5, 
               "M2" = model5a, 
               "M3" = model5b,
               "M4" = model6, 
               "M5" = model6a,
               "M6" = model6b,
               "M7" = model5c)
  
# additionally we want to change the font, font size and spacing
modelsummary(models2,
output = 'gt',
coef_map = c("index_standardized" = "Deprivation (Standarized ISDE score)",
             "classificationRural" = "Urbanicity (Rural)",
             "classificationUrbana" = "Urbanicity (Urban)",
            "zonanorte" = "Zone (North)",
            "zonasur" = "Zone (South)"),
#stars = T,
#estimate = "{estimate}{stars}",
statistic = c("SE = {std.error}", 
                "p = {p.value}{stars}"),
notes = "Notes: + p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001") %>%
tab_spanner(label = 'Dependent variable: DRSC class', columns = 2:8) %>%
tab_options(
table.font.size = 10,
data_row.padding = px(1),
table.border.top.color = "white",
heading.border.bottom.color = "black",
row_group.border.top.color = "black",
row_group.border.bottom.color = "white",
table.border.bottom.color = "white",
column_labels.border.top.color = "black",
column_labels.border.bottom.color = "black",
table_body.border.bottom.color = "black",
table_body.hlines.color = "white"
) %>% 
  tab_row_group(
    group = "",
    rows = 1:15
  )




```

```{r}
#| label: tbl-10
#| tbl-cap: "Logistic regression 2011-2023"
#| fig-align: left
#| results: "asis"


# Model 1: Single predictor (index_standardized)
model1 <- glm(class_membership ~ index_standardized, family = "binomial", data = d1)
# Model 1a: Single predictor (classification)
model1a <- glm(class_membership ~ classification, family = "binomial", data = d1)
# Model 1b: Single predictor (zona)
model1b <- glm(class_membership ~ zona, family = "binomial", data = d1)
# Model 1c: All predictors
model1c <- glm(class_membership ~ index_standardized + classification + zona, family = "binomial", data = d1)
# Model 2: Two predictors (index_standardized + classification)
model2 <- glm(class_membership ~ index_standardized + classification, family = "binomial", data = d1)
# Model 2a: Two predictors (index_standardized + zona)
model2a <- glm(class_membership ~ index_standardized + zona, family = "binomial", data = d1)
# Model 2b: Two predictors (classification + zona)
model2b <- glm(class_membership ~ classification + zona, family = "binomial", data = d1)

models <- list("M1" = model1, 
               "M2" = model1a, 
               "M3" = model1b,
               "M4" = model2, 
               "M5" = model2a,
               "M6" = model2b,
               "M7" = model1c)
  
# additionally we want to change the font, font size and spacing
modelsummary(models,
output = 'gt',
coef_map = c("index_standardized" = "Deprivation (Standarized ISDE score)",
             "classificationRural" = "Urbanicity (Rural)",
             "classificationUrbana" = "Urbanicity (Urban)",
            "zonanorte" = "Zona (Norte)",
            "zonasur" = "Zona (Sur)"),
#stars = T,
#estimate = "{estimate}{stars}",
statistic = c("SE = {std.error}", 
                "p = {p.value}{stars}"),
notes = "Notes: + p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001") %>%
tab_spanner(label = 'Dependent variable: DRSC class', columns = 2:8) %>%
tab_options(
table.font.size = 10,
data_row.padding = px(1),
table.border.top.color = "white",
heading.border.bottom.color = "black",
row_group.border.top.color = "black",
row_group.border.bottom.color = "white",
table.border.bottom.color = "white",
column_labels.border.top.color = "black",
column_labels.border.bottom.color = "black",
table_body.border.bottom.color = "black",
table_body.hlines.color = "white"
) %>% 
  tab_row_group(
    group = "",
    rows = 1:15
  )





```

```{r}
#| label: tbl-11
#| tbl-cap: "Reporting of latent growth modeling approach using the GRoLTS checklist"
#| fig-align: left


grolts <- read.csv("Grolts_report1.csv", sep=",") 


grolts %>% 
  dplyr::rename(Items = X) %>% 
  gt::gt() %>%
  #tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white") 

```

```{r}
#| label: tbl-12
#| tbl-cap: "Framework of eight steps to construct a latent class trajectory model"
#| fig-align: left

tibble::tibble(
  Step = 1:8,
  `Step description` = c(
    "Scope model by provisionally selecting a plausible number of classes based on available literature and the structure based on plausible clinical patterns.",
    "Refine the model from step 1 to confirm the optimal number of classes, typically testing K=1–7 classes.",
    "Refine optimal model structure from fixed through to unrestricted random effects of the model using the favoured K derived in step 2.",
    "Run model adequacy assessments as described in online supplementary table S3 including posterior probability of assignments (APPA), odds of correct classification (OCC) and relative entropy.",
    "Investigate graphical presentation",
    "Run additional tools to assess discrimination including Degrees of separation (DoS) and Elsensohn’s envelope of residuals",
    "Assess for clinical characterisation and plausibility.",
    "Conduct sensitivity analyses, for example, testing models without complete data at all time points."
  ),
  `Criteria for selection` = c(
    "Examine linearity of the shape of standardised residual plots for each of the classes in a model with no random effects.",
    "Lowest Bayesian information criteria value.",
    "",
    "APPA: average of maximum probabilities should be greater than 70% for all classes.\nOCC values greater than 5.0.\nRelative entropy values greater than 0.5.",
    "Plot mean trajectories across time for each class in a single graph.\nPlot mean trajectories with 95% predictive intervals for each class (one class per graph).\nPlot individual class ‘spaghetti plots’ across time for a random sample.",
    "DoS greater than zero.\nEnvelope of residuals is assessed in plots by observing clear separations between classes.",
    "Tabulation of characteristics by latent classes. Are the trajectory patterns clinically meaningful? Perhaps, consider classes with a minimum percentage of the population.\nAre the trajectory patterns clinically plausible?\nConcordance of class characteristics with those for other well-established variables.",
    "General assessment of patterns of trajectories compared with main model."
  )
) %>%
  gt::gt() %>%
  #tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white") 
```
